The impacts of demography on linked selection in maize

### 2015-07-20 manuscript.tex
update

\documentclass{pnastwo}
\usepackage{cite}
\usepackage{color}
\usepackage{pgf}
%\usepackage{hyperref}
\usepackage[normalem]{ulem}
\newcommand{\jri}[1]{\textcolor{red}{\emph{#1}} }
\newcommand{\citex}{\textcolor{red}{\bf CITE}}
\newcommand{\X}{\textcolor{red}{\bf X}}

@@

\dropcap{D}omesticated plant species evolve in a unique fashion
compared to their wild relatives \textcolor{red}{(ref here)}. This
is a result of both the anthropomorphic nature of artificial selection on
domesticates \cite{purugganan2009} as well as the demographic characteristics of the domestication
bottleneck(s) that they tend to have experienced
\cite{ross2007}. However, the
complex interplay between selective pressures and demographic
limitations, and the impact that this interplay has on identifying
selection and understanding demography, is not fully understood. Although a large body of
research that involves searching the genomes of domesticated species for evidence
of positive selection exists \cite{hufford2012, he2011, OTHERS?}, these studies tend to focus on
identifying or mapping particular genes or regions that play an
important role in phenotypic evolution. In contrast, knowledge regarding the impacts that demography and
selection have on whole-genome patterns of genetic variability therefore remains limited.

@@

\section{Results}
\subsection{Patterns of variability differ between genic and
  nongenic regions of the genome}
Previous research of the maize domestication process has relied upon
observations drawn from genic DNA \textcolor{red}{(several references
  here)}. Our data were generated through whole genome sequencing,
which eliminated this constraint. Importantly, we observed substantial
differences in patterns of diversity between genic and non-genic regions
of the genome for both maize and teosinte. For maize, mean pairwise
diversity ($\pi$) within genes was significantly lower than at
positions at least 5kb away from genes (0.00668 within, 0.00691 away, p$<$2e-44). The same
pattern of significantly greater diversity outside of genes was observed in
teosinte, but to a larger extent. For teosinte we observed $\pi$
within genes to be 0.0088 and at least 5kb from genes it was 0.1153
(p$\approx$0). \jri{that number is impossible, check again. off by maybe an order of magnitude!}
These observations suggest that genes are not evolving
neutrally in maize or teosinte. Instead, some form of selection is likely
reducing diversity within genes. Additionally, these observations
suggest that demographic inference, which relies on the assumption of
neutrally evolving DNA, will be more accurate if it is based on
observations from non-genic genetic material.

\subsection{Hard sweeps do not shape maize (or teosinte) diversity}
A mutation that is immediately beneficial and positively selected leaves a classical hard
sweep signature in the genome, whereby genetic diversity surrounding
that mutation is reduced as the haplotype where the mutation first
arose increases in frequency to fixation. The prevalence of such hard
sweeps was evaluated by comparing diversity
surrounding non-synonymous and synonymous substitutions in maize and
teosinte since divergence from \emph{Tripsacum}, roughly \textcolor{red}{ZZZ} years before
present \textcolor{red}{(ref here)}. For both taxa, no difference in diversity
surrounding these classes of substitutions was identified (Figure
\ref{fig:hardSweeps}). \jri{weird formatting that the first figure referenced is fig. 2. something about PNAS style and methods?}Additionally, diversity around maize substitutions not seen in teosinte was investigated. 
These sites have the potential
to correspond to recent maize sweeps, occuring after the split from
teosinte. Again, no difference was observed between diversity around
synonymous and nonsynonymous substitutions
\textcolor{red}{(Supplemental figure here)}. Together, these
observations suggest that hard sweeps are not a primary form of
selection for either maize or teosinte. \jri{add Enard citation and GERP results (supp. figures)}

@@

{\bf Outline}
\begin{enumerate}
\item Diversity differs between genes, not genes
\item No hard sweeps
\item soft sweeps? (putting it here means if we find a lot we may have to redo stuff below on the subset of genes w/o evidence of soft sweeps.)
\item purifying selection differs ($\pi$)
\item Estimating the bneck
\item purifying selection (singletons)
\item fit the B curve
\item simulations
\end{enumerate}

\subsection{Patterns of purifying and background selection \jri{we will need to define background selection and be careful about ``purifying selection'' vs ``background selection'', as technically background selection does not include very weak deleterious alleles.}are
  influenced by demographic history}
Purifying selection refers to the situation where deleterious mutations arising in a population are continuously selected against. 
When this form of selection is operating, it can serve to reduce genetic variability at linked neutral sites, a phenomenon called background selection \cite{charlesworth1993}. Purifying and background selection lead to lower diversity within genes and other functional sites relative to neutral regions \textcolor{red}{(ref here)}. 
We investigated purifying selection in maize and teosinte by evaluating the average magnitude of reduced diversity within genes and recovery away from genes in both taxa. 
When standardized by neutral levels, which were defined as mean diversity at postions at least 0.01 cM distal
from genes, a stronger reduction of diversity and slower recovery was observed for teosinte than for maize, implying that purifying selection has left a more pronounced signature in the teosinte genome (Figure \ref{fig:purify}). 
This conflicted with our \emph{a priori} hypothesis;
we expected that strong artificial selection since domestication may \jri{don't think we should say this; this wasn't my a priori prediction}
have elevated the intensity of purifying selection for maize. 
We therefore conducted a parallel analysis based on singleton diversity. \jri{disagree singletons should go here. see outline above.}
As a class, singleton alleles depict the most recent patterns of evolution, but
also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosite (n=13), maize singletons were
analyzed directly as well as downsampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was modified to be mean diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged. Maize singleton diversity was just at least as low as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{fig:purify}), implying that in the
recent past maize has been at least as influenced by purifying selection as
teosinte. Together, these findings suggest that demographic history has
a strong influence on the effect of purifying selection. Historically,
teosinte has had a larger population size than maize, and only
recently has maize population size overcome that of teosinte. Since
the efficacy of purifying selection scales with population size,
these results likely reflect changes in $N_e$ more than they reflect
underlying changes in selection pressure.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent
  past. \label{fig:purify} \jri{are these done with latest values we used for curve fitting? should also add the singleton vs pi comparison within species as a supplemental figure. if $\pi$/singleton curves {\bf don't} match, will need to explain this suggests teosinte not in equilibrium.  good justification i think for msmc model, which we then use to say ``gee whiz, not perfect equilibrium but who cares because the big interesting demography is in maize.'' }
\end{figure*}

\textcolor{red}{This might be a good place to stick in a paragraph
  about curve fitting B to estimate s, mu.}

@@

\subsection{Simulations confirm the relationship between demographic
  processes and background selection}
\textcolor{red}{We explored whether or not the demographic patterns
  that correspond to maize domestication are capable of generating the
  interesting pattern of background selection that is observed. That
  is, does demography alone explain the weak effect of background
  selection on pairwise diversity in maize? We begain by estimating
  the distribution of fitness effects and mutation rate in both maize and teosinte from
  (formulas)... These were observed to be ZZZZZZ. Then, we simulated
  mating according to our estimated demographic model and evaluated
  the resulting levels of pairwise and singleton diversity. A LOT MORE
GOES HERE, BUT WHAT IS WRITTEN NOW CAN FRAME THINGS.}

@@

\subsection{Plant materials}
Accessions studied were selected from the Maize HapMap2
panel \cite{chia2012}. \jri{aren't you using the high coverage teosinte? if so I think you need to cite Lemmon 2014 Plos Gen} 
Principal component analysis was employed to ensure that closely related individuals were not included due to their potential to bias results \textcolor{red}{(maybe a supplemental figure here)}. \jri{yes supp. figure good idea}
Ultimately, 23 maize inbreds derived from a diverse assortment of landraces were selected for inclusion. 
Thirteen teosinte inbred lines, all members of  subspecies Z. \emph{mays}
ssp. \emph{parviglumis}, were utilized. 
Sequences were mapped to the maize B73 version 3 reference genome \cite{schnable2009} \url(ftp://ftp.ensemblgenomes.org/pub/plants/release-22/fasta/zea\_mays/dna/).

@@

% \begin{thebibliography}{10}
% \bibitem{dadi}
% R.~Gutenkunst, R.~Hernandez,S.~Williamson, and C.~Bustamante, {\em
%   Inferring the joint demographic history of multiple populations from
%   multidimensional SNP frequency data}, PLoS Genetics., 5:10 (2009), e1000695.

% \bibitem{hapmap2}
% J.~Chia, C.~Song, P.~Bradbury, D.~Costich, N.~de~Leon, and others,
% \emph{Maize HapMap2 identifies extant variation from a genome in
%   flux}, Nature Genetics., 44 (2012), 803-807.

% \bibitem{maizeGenome}
% P.~Schnable, D.~Ware, R.~Fulton, J.~Stein, F.~Wei, \emph{The B73 maize
% genome:complexity, diversity, and dynamics}, Science., 326:5956
% (2009), 1112-1115.

% \bibitem{angsd}
% T.~Korneliussen, A.~Albrechtsen, and R.~Nielsen, \emph{ANGSD: Analysis of
% next generation sequencing data}, BMC Bioinformatics., 15:356 (2014),
% 10.1186/s12859-014-0356-4.

% \bibitem{R}
% R Core Team, \emph{R: A language and environment for statistical
%   computing}, R Foundation for Statistical Computing., Vienna,
% Austria.

% \bibitem{biomaRt1}
% S.~Durinck, P.T.~Soekknabm E.~Birney, and W.~Huber, \emph{Mapping
%   identifiers for the integration of genomic datasets with the
%   R/Bioconductor package biomaRt}, Nature Protocols., 4 (2009),
% 1184-1191.

% \bibitem{biomaRt2}
% S.~Durinck, Y.~Moreau, A.~Kasprzyk, S.~Davis, B.~De~Moor, and others,
% \emph{BioMart and Bioconductor: a powerful link between biological
%   databases and microarray data analysis}, Bioinformatics, 21 (2005),
% 3439-3440.

% \bibitem{vep}
% W.~McLaren, B.~Pritchard, D.~Rios, Y.~Chen, P.~Flicek, and
% F.~Cunningham, \emph{Deriving the consequences of genomic variants
%   with the Ensembl API and SNP Effect Predictor}, Bioinformatics,
% 26:16 (2010), 2069-2070.

% \bibitem{hernandez11}
% R.D.~Hernandez, J.L.~Kelley, E.~Elyashiv, S.C.~Melton, A.~Auton, and
% others, \emph{Classic selective sweeps were rare in recent human
%   evolution}, Science, 331:6019 (2011), 920-924.

% \bibitem{sattath11}
% S.~Sattath, E.~Elyashiv, O.~Kolodny, Y.~Rinott, and G.~Sella,
% \emph{Pervasive adaptative protein evolution apparent in diversity
%   patterns around amino acid substitutions in Drosophila simulans},
% PLoS Genetics, 7:2 (2011), e1001302.

% \bibitem{charlesworth93}
% B.~Charlesworth, M.~T.~Morgan, and D.~Charlesworth, \emph{The Effect
%   of Deleterious Mutations on Neutral Molecular Variation}, Genetics,
% 134 (1993), 1289-1303.

% \end{thebibliography}

>>>>

\documentclass{pnastwo}
\usepackage{cite}
\usepackage{color}
\usepackage{pgf}
%\usepackage{hyperref}
\usepackage[normalem]{ulem}
\newcommand{\jri}[1]{\textcolor{blue}{\emph{#1}} }
\newcommand{\citex}{\textcolor{blue}{\bf CITE}}
\newcommand{\X}{\textcolor{blue}{\bf X}}

@@

\dropcap{D}omesticated plant species evolve in a unique fashion
compared to their wild counterparts \textcolor{red}{(ref here)}. This
is a result of both the anthropomorphic nature of artificial selection on
domesticates \cite{purugganan2009} as well as the demographic characteristics of the domestication
bottleneck(s) that they tend to have experienced
\cite{ross2007}. However, the
complex interplay between selective pressures and demographic
limitations, and the impact that this interplay has on identifying
selection and understanding demography, is not fully understood. Although a large body of
research that involves searching the genomes of domesticated species for evidence
of positive selection exists \cite{hufford2012, he2011, OTHERS?}, these studies tend to focus on
identifying or mapping particular genes or regions that play an
important role in phenotypic evolution. In contrast, knowledge regarding the impacts that demography and
selection have on whole-genome patterns of genetic variability therefore remains limited.

@@

\section{Results}
\subsection{Patterns of diversity differ between genic and
  nongenic regions of the genome}
Previous research of the maize domestication process has relied upon
observations drawn from genic DNA \textcolor{red}{(several references
  here)}. Our data were generated through whole genome sequencing,
which eliminated this constraint. Importantly, we observed substantial
differences in patterns of diversity between genic and non-genic regions
of the genome for both maize and teosinte. For maize, mean pairwise
diversity ($\pi$) within genes was significantly lower than at
positions at least 5kb away from genes (0.00668 within, 0.00691 away, p$<$2e-44). The same
pattern of significantly greater diversity outside of genes was observed in
teosinte, but to a larger extent. For teosinte we observed $\pi$
within genes to be 0.0088 and at least 5kb from genes it was 0.0115
(p$\approx$0). These observations suggest that genes are not evolving
neutrally in maize or teosinte. Instead, some form of selection is likely
reducing diversity within genes. Additionally, these observations
suggest that demographic inference, which relies on the assumption of
neutrally evolving DNA, will be more accurate if it is based on
observations from non-genic genetic material.

\subsection{Hard sweeps do not shape maize (or teosinte) diversity}
A mutation that is immediately beneficial and positively selected leaves a classical hard
sweep signature in the genome, whereby genetic diversity surrounding
that mutation is reduced as the haplotype where the mutation first
arose increases in frequency to fixation. The prevalence of such hard
sweeps was evaluated by comparing diversity
surrounding non-synonymous and synonymous substitutions in maize and
teosinte since divergence from \emph{Tripsacum}, roughly 1-1.2 million years before
present \cite{ross2009}. For both taxa, no difference in diversity
surrounding these classes of substitutions was identified (Figure
\ref{fig:hardSweeps}). This observation suggests that hard sweeps are not a primary form of
selection for either maize or teosinte.

To ensure that this analysis is resilient to the intricacies of our sample, we investigated diversity surrounding the maize gene tga1, one of the few known instances of a hard sweep corresponding to a causal nonsynonymous mutuation in maize \cite{wang2015}. In the case of this known sweep, we observed a reduction in diversity surrounding the causitive substitution to levels much lower than seen for synonymous substitutions \textcolor{red}{(SUPP FIG)}, confirming that this method should be informative if hard sweeps were the prevaling mode of selection in maize. Still, to ensure that the approach is robust to potential shortcomings such as limited power in the face of background selection \cite{enard2014}, we sought to determine if patterns of diversity surrounding synonymous and non-synonymous substitutions differ when we restrict our analysis to the most- or least-conserved genes in maize. When we re-analyzed subsets of genes with the 10\% highest and 10\% lowest genomic evolutionary rate profile (GERP) score \cite{davydov2010, cooper2005, rodgers2015}, we again saw no significant difference in diversity surrounding synonymous and non-synonymous sites \textcolor{red}{(SUPP FIG)}.

\subsection{Soft Sweeps?}
In this section, I'll write about what we did to look for soft sweeps...

@@

%%%%
%%%% This outline no longer needed in-text because results headers follow
%%%%
% \jri{
% {\bf Outline}
% \begin{enumerate}
% \item Diversity differs between genes, not genes
% \item No hard sweeps
% \item soft sweeps? (putting it here means if we find a lot we may have to redo stuff below on the subset of genes w/o evidence of soft sweeps.)
% \item purifying selection differs ($\pi$)
% \item Estimating the bneck
% \item purifying selection (singletons)
% \item fit the B curve
% \item simulations
% \end{enumerate}
% }

\subsection{Patterns of purifying and background selection are
  influenced by demographic history}
 \jri{we will need to define background selection and be careful about ``purifying selection'' vs ``background selection'', as technically background selection does not include very weak deleterious alleles.}
Purifying selection refers to the situation where deleterious mutations arising in a population are continuously selected against.
When this form of selection is operating, it can serve to reduce genetic variability at linked neutral sites, a phenomenon called background selection \cite{charlesworth1993}. Purifying and background selection lead to lower diversity within genes and other functional sites relative to neutral regions \textcolor{red}{(ref here)}.
We investigated purifying selection in maize and teosinte by evaluating the average magnitude of reduced diversity within genes and recovery away from genes in both taxa.
When standardized by neutral levels, which were defined as mean diversity at postions at least 0.01 cM distal
from genes, a stronger reduction of diversity and slower recovery was observed for teosinte than for maize, implying that purifying selection has left a more pronounced signature in the teosinte genome (Figure \ref{fig:purify}).
This conflicted with our \emph{a priori} hypothesis;
we expected that strong artificial selection since domestication may \jri{don't think we should say this; this wasn't my a priori prediction}
have elevated the intensity of purifying selection for maize.
We therefore conducted a parallel analysis based on singleton diversity. \jri{disagree singletons should go here. see outline above.}
As a class, singleton alleles depict the most recent patterns of evolution, but
also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosite (n=13), maize singletons were
analyzed directly as well as downsampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was modified to be mean diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged. Maize singleton diversity was just at least as low as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{fig:purify}), implying that in the
recent past maize has been at least as influenced by purifying selection as
teosinte. Together, these findings suggest that demographic history has
a strong influence on the effect of purifying selection. Historically,
teosinte has had a larger population size than maize, and only
recently has maize population size overcome that of teosinte. Since
the efficacy of purifying selection scales with population size,
these results likely reflect changes in $N_e$ more than they reflect
underlying changes in selection pressure.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent
  past. \label{fig:purify} \jri{are these done with latest values we used for curve fitting? should also add the singleton vs pi comparison within species as a supplemental figure. if $\pi$/singleton curves {\bf don't} match, will need to explain this suggests teosinte not in equilibrium.  good justification i think for msmc model, which we then use to say ``gee whiz, not perfect equilibrium but who cares because the big interesting demography is in maize.'' }}
\end{figure*}

@@

\section{purifying selection (singletons)}
If this is separated out from the previous purifying selection section
and placed after the demography section, it should go here.

\subsection{Simulations confirm the relationship between demographic
  processes and background selection}
\textcolor{red}{This might be a good place to stick in a paragraph
  about curve fitting B to estimate s, mu.}

\textcolor{red}{We explored whether or not the demographic patterns
  that correspond to maize domestication are capable of generating the
  interesting pattern of background selection that is observed. That
  is, does demography alone explain the weak effect of background
  selection on pairwise diversity in maize? We begain by estimating
  the distribution of fitness effects and mutation rate in both maize and teosinte from
  (formulas)... These were observed to be ZZZZZZ. Then, we simulated
  mating according to our estimated demographic model and evaluated
  the resulting levels of pairwise and singleton diversity. A LOT MORE
GOES HERE, BUT WHAT IS WRITTEN NOW CAN FRAME THINGS.}

@@

\subsection{Plant materials}
Accessions studied were selected from the Maize HapMap2
panel \cite{chia2012, lemmon2014}. Principal component analysis was employed to ensure that closely related individuals were not included due to their potential to bias results \textcolor{red}{(maybe a supplemental figure here)}. \jri{yes supp. figure good idea}
Ultimately, 23 maize inbreds derived from a diverse assortment of landraces were selected for inclusion.
Thirteen teosinte inbred lines, all members of  subspecies Z. \emph{mays}
ssp. \emph{parviglumis}, were utilized.
Sequences were mapped to the maize B73 version 3 reference genome \cite{schnable2009} \url(ftp://ftp.ensemblgenomes.org/pub/plants/release-22/fasta/zea\_mays/dna/).

@@
